---
title: "Exploring economics and quality of life"
author: "Tim Szewczyk"
date: "2017 April 28"
output:
  pdf_document: default
  html_notebook: default
subtitle: The Data Incubator Capstone Project
---

# Overview
This project creates a tool to explore patterns between economic/demographic indicators, quality of life, and voting patterns. All data are at the county level. Ideas for the main output:  

* An interactive map that allows users to choose a state(s) and compare data for the counties within the state(s).  
* An interactive scatter plot to compare particular variables, with the selected state(s) or county(ies) highlighted in the plot.  
* A hierarchical model using latent variables to describe quality of life, access to public goods, etc.  
* A hierarchical model predicting 2016 election results (& change from 2008, 2012) based on these variables.  
* Other things  

The data sources I intend to use are:  

* 2010 census data  
* Health data from the JUST Capital posting  
* USDA data on food quality & prices  
* Museum data  
* Library data  
* County-level elections results  
* Cost of Living index in the future?  

First, I just need to get the data all organized and cleaned, and then make a couple of interesting plots to show that this could actually go somewhere.  


# Data cleaning  
```{r setEnv, message=FALSE}
## load libraries
library(tidyverse); theme_set(theme_bw()); library(sevcheck)

## read in data
# FIPS reference for identifying counties
fips.i <- read.csv("data/fips_codes_website.csv")
fips.i$FIPS <- paste0(sprintf("%02d", fips.i$State.FIPS.Code),
                      sprintf("%03d", fips.i$County.FIPS.Code))
# census data
cens.df <- read.csv("data/CENSUS_DATA.csv")
cens.df <- cens.df[, -c(grep("CI90", names(cens.df)), 
                        grep("Estimate", names(cens.df)))]
cens.df$FIPS <- sprintf("%05d", cens.df$FIPS)
# personal income data
incm.df <- read.csv("data/lapi1116.csv")
incm.df$FIPS <- fips.i$FIPS[match(paste(incm.df$State,
                                        incm.df$County, sep="."),
                                  paste(fips.i$State.Abbreviation,
                                        fips.i$GU.Name, sep="."))]
# health data
hlth.df <- read.csv("data/HEALTH_DATA.csv")
hlth.df <- hlth.df[, c(1:3, sort(c(grep("Pct", names(hlth.df)), 
                                   grep("Rate", names(hlth.df)),
                                   grep("Rank", names(hlth.df)))))]
hlth.df$FIPS <- sprintf("%05d", hlth.df$FIPS)
# museum data
musm.df <- read.csv("data/Museum_Universe_Data_File_FY_2015_Q3.csv")
musm.df$FIPS <- paste0(sprintf("%02d", musm.df$FIPSST),
                      sprintf("%03d", musm.df$FIPSCO))
# library data
libr.df <- read.csv("data/PLS_FY2014_AE_pupld14a.csv")
libr.df$FIPS <- paste0(sprintf("%02d", libr.df$FIPSST),
                      sprintf("%03d", libr.df$FIPSCO))
# election data
elec.df <- read.csv("data/US_County_Level_Presidential_Results_08-16.csv")
elec.df$FIPS <- sprintf("%05d", elec.df$fips_code)
elec.df[,16:24] <- NA
names(elec.df)[16:24] <- paste0("VS_", 
                                rep(c("D_", "R_", "O_"), times=3), 
                                rep(c(2008, 2012, 2016), each=3))
elec.df[,16:18] <- elec.df[,4:6]/elec.df[,3]
elec.df[,19:21] <- elec.df[,8:10]/elec.df[,7]
elec.df[,22:24] <- elec.df[,12:14]/elec.df[,11]
# calculate county summaries
elec.df$CO_2008 <- ifelse(elec.df$dem_2008 > elec.df$gop_2008, "D", "R")
elec.df$CO_2012 <- ifelse(elec.df$dem_2012 > elec.df$gop_2012, "D", "R")
elec.df$CO_2016 <- ifelse(elec.df$dem_2016 > elec.df$gop_2016, "D", "R")
# calculate state summaries
elec.df$State <- fips.i$State.Abbreviation[match(elec.df$FIPS, fips.i$FIPS)]
elSt.df <- aggregate(elec.df[,c(3:14)], 
                     by=list(State=elec.df$State), FUN=sum)
elSt.df$ST_2008 <- ifelse(elSt.df$dem_2008 > elSt.df$gop_2008, "D", "R")
elSt.df$ST_2012 <- ifelse(elSt.df$dem_2012 > elSt.df$gop_2012, "D", "R")
elSt.df$ST_2016 <- ifelse(elSt.df$dem_2016 > elSt.df$gop_2016, "D", "R")
elec.df$ST_2008 <- elSt.df$ST_2008[match(elec.df$State, elSt.df$State)]
elec.df$ST_2012 <- elSt.df$ST_2012[match(elec.df$State, elSt.df$State)]
elec.df$ST_2016 <- elSt.df$ST_2016[match(elec.df$State, elSt.df$State)]
elec.df$ST_SWNG <- elSt.df$ST_2016[match(elec.df$State, elSt.df$State)]
swng <- c("WI", "PA", "NH", "MN", "OH", "IA", "VA", "FL", "MI", "NV", "CO", "NC")
elec.df$ST_SWNG[elec.df$State %in% swng] <- "S"
elec.df$ST_SWNG <- factor(elec.df$ST_SWNG, levels=c("D", "S", "R"),
                          labels=c("Democratic", "Swing", "Republican"))
elec.df$VS_D_12_16 <- elec.df$VS_D_2016 - elec.df$VS_D_2012
elec.df$VS_R_12_16 <- elec.df$VS_R_2016 - elec.df$VS_R_2012
# urbanization code
urbn.df <- read.csv("data/NCHSURCodes2013.csv")
urbn.df$FIPS <- sprintf("%05d", urbn.df$fips_code)
urbn.df$urb.code2013 <- factor(urbn.df$urb.code2013, 
                               labels=c("Large central metro", 
                                        "Large fringe metro",
                                        "Medium metro",
                                        "Small metro",
                                        "Micropolitan",
                                        "Noncore"))

## merge data
all.df <- merge(cens.df[,c(3,6:9)], incm.df[,c(3:10)], by="FIPS") %>%
  merge(hlth.df[,c(1,4:77)], by="FIPS") %>%
  merge(elec.df[,c(3:34)], by="FIPS") %>%
  merge(urbn.df[,c(6,7,10)])
```

# EDA  
The main point of this project is exploration. At least for now. 
```{r eda}
# food insecurity in D vs R states
ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=logit(VS_D_2008))) + facet_wrap(~ST_2008) +
  geom_hline(yintercept=0, linetype=3) + geom_point(alpha=0.5)
ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=logit(VS_D_2012))) + facet_wrap(~ST_2012) +
  geom_hline(yintercept=0, linetype=3) + geom_point(alpha=0.5)
ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=logit(VS_D_2016))) + facet_wrap(~ST_2016) +
  geom_hline(yintercept=0, linetype=3) + geom_point(alpha=0.5)

ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=log(dem_2008))) +
  geom_point(alpha=0.5) + facet_wrap(~ST_2008)
ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=log(dem_2012))) +
  geom_point(alpha=0.5) + facet_wrap(~ST_2012)
ggplot(all.df, aes(x=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   y=log(dem_2016))) +
  geom_point(alpha=0.5) + facet_wrap(~ST_2016)

ggplot(all.df, aes(y=logit(Food.insecurity.Pct.Food.Insecure/100), 
                   x=log(Median.Household.Income), colour=ST_2016)) +
  geom_point(alpha=0.5) + scale_colour_manual(values=c("blue", "red")) + 
  stat_smooth(method="lm")
lm(logit(Food.insecurity.Pct.Food.Insecure/100) ~ log(Median.Household.Income)*ST_2016,
   data=all.df) %>% summary 

# population vs voting
ggplot(all.df, aes(x=log(total_2008), colour=CO_2008)) + geom_density() +
  scale_colour_manual(values=c("blue", "red"))

ggplot(all.df, aes(x=Demographics.Pct.Non.Hispanic.White, y=Food.insecurity.Pct.Food.Insecure)) + geom_point(alpha=0.5) + facet_wrap(~ST_2012) + labs(x="% White", y="% Food Insecure")


ggplot(all.df, aes(x=urb.code2013, y=logit(VS_D_2012))) + geom_boxplot()

```


```{r storingPlots, warning=FALSE}
# makes and stores a whole bunch of graphs
# plot parameters
swng.cols <- c("blue", "purple3", "firebrick2")
dr.cols <- swng.cols[c(1,3)]
swng.dim <- c("height"=4, "width"=11)
urb.dim <- c("height"=6, "width"=10)

# transformations
dat.col <- c(2:8,10:11,23:86)
all.sc <- all.df
log.col <- c(5:8,23,32:33,35:38,50:51,58:59,63,66:67,71,74:75)
logit.col <- c(2:4,24:31,34,39:49,52:57,60:62,64:65,68:70,72:73,77:86)
all.sc[,log.col] <- log(all.sc[,log.col])
all.sc[,logit.col] <- logit(all.sc[,logit.col]/100)

# plotting loop
for(i in dat.col) {
  ##### variable histograms
  jpeg(paste0("EQ_EDA/hist/", i, ".jpg"))
    hist(all.df[,i], breaks=30, main=names(all.df)[i])
  dev.off()
  jpeg(paste0("EQ_EDA/hist/", i, "_sc.jpg"))
    hist(all.sc[,i], breaks=30, main=paste(names(all.sc)[i], "Scaled"))
  dev.off()

  
  ##### county parties
  co.party.plot <- ggplot(all.sc, aes_string(y=names(all.df)[i], 
                                             x="VS_D_2008",
                                    colour="CO_2008")) +
    geom_vline(xintercept=0.5, linetype=3) +
    geom_point(shape=1, alpha=0.5) + facet_grid(~ST_SWNG) +
    stat_smooth(method="loess", span=1, colour="gray30", se=F) +
    stat_smooth(data=filter(all.sc, CO_2008=="R"), se=F, size=0.5,
                method="loess", span=1, colour="firebrick4") +
    stat_smooth(data=filter(all.sc, CO_2008=="D"), se=F, size=0.5,
                method="loess", span=1, colour="blue3") +
    geom_rug(sides="lb", alpha=0.07) + labs(y=names(all.df)[i]) +
    scale_colour_manual(values=dr.cols) + xlim(0,1)
  # 2008
  gg.p <- co.party.plot + labs(x="% Dem 2008")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2008D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2008") + labs(x="% Rep 2008")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2008R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2012
  gg.p <- co.party.plot %+% 
    aes_string(x="VS_D_2012", colour="CO_2012") + labs(x="% Dem 2012")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2012D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2012") + labs(x="% Rep 2012")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2012R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2016
  gg.p <- co.party.plot %+% 
    aes_string(x="VS_D_2016", colour="CO_2016") + labs(x="% Dem 2016")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2016D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2016") + labs(x="% Rep 2016")
  ggsave(paste0("EQ_EDA/CO_Party/", names(all.df)[i], "_2016R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2016 - 2012
  gg.p <- co.party.plot %+% aes_string(x="VS_D_12_16", colour="CO_2016") + 
    geom_vline(xintercept=0, linetype=3) + 
    labs(x="(% Dem 2016) - (% Dem 2012)") + xlim(-0.25,0.25)
  ggsave(paste0("EQ_EDA/CO_Party/chg/", names(all.df)[i], "_16-12D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_12_16") + 
    labs(x="(% Rep 2016) - (% Rep 2012)")
  ggsave(paste0("EQ_EDA/CO_Party/chg/", names(all.df)[i], "_16-12R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  
  
  ##### urbanization (facet by swing status)
  urb.plot <- ggplot(all.sc, aes_string(y=names(all.df)[i], 
                                        x="VS_D_2008",
                                    colour="urb.code2013")) +
    geom_vline(xintercept=0.5, linetype=3) +
    geom_point(shape=1, alpha=0.8) + facet_grid(~ST_SWNG) +
    stat_smooth(method="loess", span=1, colour="gray30", se=F) +
    stat_smooth(se=F, size=0.5, method="loess", span=1) +
    geom_rug(sides="lb", size=0.1) +
    scale_colour_brewer(type="div") +
    labs(y=names(all.df)[i]) + xlim(0,1)
  # 2008
  gg.p <- urb.plot + labs(x="% Dem 2008")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2008D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2008") + labs(x="% Rep 2008")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2008R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2012
  gg.p <- urb.plot %+% aes_string(x="VS_D_2012") + labs(x="% Dem 2012")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2012D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2012") + labs(x="% Rep 2012")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2012R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2016
  gg.p <- urb.plot %+% aes_string(x="VS_D_2016") + labs(x="% Dem 2016")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2016D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2016") + labs(x="% Rep 2016")
  ggsave(paste0("EQ_EDA/Urban_bySwng/", names(all.df)[i], "_2016R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  # 2016 - 2012
  gg.p <- urb.plot %+% aes_string(x="VS_D_12_16") + 
    geom_vline(xintercept=0, linetype=3) + 
    labs(x="(% Dem 2016) - (% Dem 2012)") + xlim(-0.25,0.25)
  ggsave(paste0("EQ_EDA/Urban_bySwng/chg/", names(all.df)[i], "_16-12D.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_12_16") + labs(x="(% Rep 2016) - (% Rep 2012)")
  ggsave(paste0("EQ_EDA/Urban_bySwng/chg/", names(all.df)[i], "_16-12R.jpg"), gg.p,
         height=swng.dim[1], width=swng.dim[2])
  
  
  ##### urbanization (facet by urban code)
  urb.plot2 <- ggplot(all.sc, aes_string(y=names(all.df)[i], 
                                         x="VS_D_2008",
                                         colour="ST_SWNG")) +
    geom_vline(xintercept=0.5, linetype=3) +
    geom_point(shape=1, alpha=0.7) + facet_wrap(~urb.code2013) +
    stat_smooth(method="loess", span=1, colour="gray30", se=F) +
    stat_smooth(se=F, size=0.5, method="loess", span=1) +
    geom_rug(sides="lb", size=0.1) +
    scale_colour_manual(values=swng.cols) +
    labs(y=names(all.df)[i]) + xlim(0,1)
  # 2008
  gg.p <- urb.plot2 + labs(x="% Dem 2008")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2008D.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2008") + labs(x="% Rep 2008")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2008R.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  # 2012
  gg.p <- urb.plot2 %+% aes_string(x="VS_D_2012") + labs(x="% Dem 2012")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2012D.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2012") + labs(x="% Rep 2012")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2012R.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  # 2016
  gg.p <- urb.plot2 %+% aes_string(x="VS_D_2016") + labs(x="% Dem 2016")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2016D.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_2016") + labs(x="% Rep 2016")
  ggsave(paste0("EQ_EDA/Urban_byUrb/", names(all.df)[i], "_2016R.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  # 2016 - 2012
  gg.p <- urb.plot2 %+% aes_string(x="VS_D_12_16") + 
    geom_vline(xintercept=0, linetype=3) + 
    labs(x="(% Dem 2016) - (% Dem 2012)") + xlim(-0.25,0.25)
  ggsave(paste0("EQ_EDA/Urban_byUrb/chg/", names(all.df)[i], "_16-12D.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
  gg.p <- gg.p %+% aes_string(x="VS_R_12_16") + labs(x="(% Rep 2016) - (% Rep 2012)")
  ggsave(paste0("EQ_EDA/Urban_byUrb/chg/", names(all.df)[i], "_16-12R.jpg"), gg.p,
         height=urb.dim[1], width=urb.dim[2])
}





```







